---
title: The Hitchikers Guide to Inferential Statistics
author: ''
date: '2020-07-17'
slug: hitchikers-guide-to-inferential-statistics
categories: ['hitchikers_guide']
tags: []
keywords:
  - tech
---

<!--more-->

```{r, echo = FALSE, warning = FALSE, message = FALSE}
library(ggplot2)
library(ggthemes)
library(dplyr)
library(stringr)
theme_set(theme_economist())

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```


# Inferential vs Descriptive stats.

- Descriptive - Purely used to summarize data, like IQR, Mean. Importantly, they do not draw conclusions beyond the data we already have.
- Inferential - Does allow us to make conclusions beyond the data we have.

## Hypothesis Testing

Inferential statistics is based on the premise that you can't prove something is true, but you can disprove something by finding an exception.

Decide what you are trying to find evidence for (alternative hypothesis), then set up the opposite as null hypothesis, and find evidence to disprove that. 

Suppose you want to show a link between studying and test results. 


- $H_0$: There is no effect on test result from studying
- $H_1$: There is an effect on test result from studying


Hypotheses are always about the population parameters (like the mean).

So the above should be stated mathematically:


- $H_0$: $\mu_0$ = $\mu_1$
- $H_1$: $\mu_0$ =/= $\mu_1$

If $\mu_0$ and $\mu_1$ are the means for the "no study" and "study" groups respectively.

General formula for test statistics: $\frac{(Observed Data) - (What We Expect if the null is true)}{AverageVariation}$.
This formula gets adapted for all the specific cases.


## Standardization

(Example from [here](https://www.youtube.com/watch?v=uAxyI_XfqXk&list=PL8dPuuaLjXtNM_Y-bUAhblSAdWRnmBUcr&index=19))

How can we compare things that aren't the same?

Suppose there are two tests: SAT and ACT.

Both try measure college readiness, but SAT is out of 1600 points and ACT is out of 36. How can you compare scores across these 2 tests?

 - Centre both distributions around 0 (subtract the mean) Mean of SAT is 1000, and mean of ACT is 21, so subtract those.
 - Scales are still wrong, so divide adjusted score by standard deviation


This adjusted score is called a Z-Score (see Definitions). $Z = \frac{x - \mu}{\sigma}$.

Note the assumption that both the distribution of SAT and ACT are normally distributed!

## Normal Distribution

(Based on [here](https://www.youtube.com/watch?v=rBjft49MAO8&list=PL8dPuuaLjXtNM_Y-bUAhblSAdWRnmBUcr&index=20))

Normal distribution is important not only because a lot of things are normally distributed, like, height, IQ etc...
There are a lot of things like debt, blood pressure which are not normally distributed. But here's the thing - distributions of means are normally distributed, even if populations aren't. So, if you sample a lot from a population which is not distributed normally, the mean of those samples will be normally distributed!


In order to meaningfully compare whether two means are different, we need to know something about their distribution: the sampling distribution.


This is described formally in the __Central Limit Theorem__ (see definitions). 


```{r}
#roll, as in roll dice
# m = number of times
# n = number of dice

roll <- function(m, n){
  means <- plyr::ldply(1:m, function(x){
    return(mean(sample(1:6, n, replace = TRUE)))
  }) 
}

n_ = 1
m_ = 10000
g <- ggplot(roll(m = m_, n = n_),
            mapping = aes(x = V1)) +
  geom_density() +
  geom_vline(xintercept = 3.5, colour = "tomato3") +
  labs(
    subtitle = str_interp('Density of means of dice rolls for ${n_} dice over ${m_} rolls.'),
    x = 'Mean Score'
  ) +
  geom_bar(aes(y = ..prop..), alpha = 0.4) +
  scale_x_continuous(
    breaks = 1:6,
    lim =  c(0, 7)
  )

g



n_ = 2
m_ = 10000
g <- ggplot(roll(m = m_, n = n_),
            mapping = aes(x = V1)) +
  geom_density() +
  geom_vline(xintercept = 3.5, colour = "tomato3") +
  labs(
    subtitle = str_interp('Density of means of dice rolls for ${n_} dice over ${m_} rolls.'),
    x = 'Mean Score'
  ) +
  geom_bar(aes(y = ..prop..), width = 0.1, alpha = 0.4) +
  scale_x_continuous(
    breaks = 1:6,
    lim =  c(0, 7)
  )


g

n_ = 3
m_ = 10000
g <- ggplot(roll(m = m_, n = n_),
            mapping = aes(x = V1)) +
  geom_density() +
  geom_vline(xintercept = 3.5, colour = "tomato3") +
  labs(
    subtitle = str_interp('Density of means of dice rolls for ${n_} dice over ${m_} rolls.'),
    x = 'Mean Score'
  ) +
  geom_bar(aes(y = ..prop..), width = 0.1, alpha = 0.4) +
  scale_x_continuous(
    breaks = 1:6,
    lim =  c(0, 7)
  )


g

n_ = 4
m_ = 10000
g <- ggplot(roll(m = m_, n = n_),
            mapping = aes(x = V1)) +
  geom_density() +
  geom_vline(xintercept = 3.5, colour = "tomato3") +
  labs(
    subtitle = str_interp('Density of means of dice rolls for ${n_} dice over ${m_} rolls.'),
    x = 'Mean Score'
  ) +
  geom_bar(aes(y = ..prop..), width = 0.1, alpha = 0.4) +
  scale_x_continuous(
    breaks = 1:6,
    lim =  c(0, 7)
  )


g

n_ = 5
m_ = 10000
g <- ggplot(roll(m = m_, n = n_),
            mapping = aes(x = V1)) +
  geom_density() +
  geom_vline(xintercept = 3.5, colour = "tomato3") +
  labs(
    subtitle = str_interp('Density of means of dice rolls for ${n_} dice over ${m_} rolls.'),
    x = 'Mean Score'
  ) +
  geom_bar(aes(y = ..prop..), width = 0.1, alpha = 0.4) +
  scale_x_continuous(
    breaks = 1:6,
    lim =  c(0, 7)
  )


g

n_ = 20
m_ = 10000
g <- ggplot(roll(m = m_, n = n_),
            mapping = aes(x = V1)) +
  geom_density() +
  geom_vline(xintercept = 3.5, colour = "tomato3") +
  labs(
    subtitle = str_interp('Density of means of dice rolls for ${n_} dice over ${m_} rolls.'),
    x = 'Mean Score'
  ) +
  geom_bar(aes(y = ..prop..), width = 0.1, alpha = 0.4) +
  scale_x_continuous(
    breaks = 1:6,
    lim =  c(0, 7)
  )

g

n_ = 200
m_ = 10000
g <- ggplot(roll(m = m_, n = n_),
            mapping = aes(x = V1)) +
  geom_density() +
  geom_vline(xintercept = 3.5, colour = "tomato3") +
  labs(
    subtitle = str_interp('Density of means of dice rolls for ${n_} dice over ${m_} rolls.'),
    x = 'Mean Score'
  ) +
  geom_bar(aes(y = ..prop..), width = 0.1, alpha = 0.4) +
  scale_x_continuous(
    breaks = 1:6,
    lim =  c(0, 7)
  )

g

n_ = 2000
m_ = 10000
g <- ggplot(roll(m = m_, n = n_),
            mapping = aes(x = V1)) +
  geom_density() +
  geom_vline(xintercept = 3.5, colour = "tomato3") +
  labs(
    subtitle = str_interp('Density of means of dice rolls for ${n_} dice over ${m_} rolls.'),
    x = 'Mean Score'
  ) +
  geom_bar(aes(y = ..prop..), width = 0.1, alpha = 0.4) +
  scale_x_continuous(
    breaks = 1:6,
    lim =  c(0, 7)
  )

g
```

Now, the mean of the distribution of sample means is the same as the population mean, its standard deviation is not.

However, the standard deviation of the distribution of sample means _is_ related to the standard deviation of the population. But, it's also related to the sample size. In the above graphs, see how the probability distributions of the graphs get thinner, as the number of dice goes up? The bigger your sample size, the closer your sample means are to the true population mean. So we need to adjust the original population standard deviation somehow to reflect this:

$\sigma_{samplingDistribution} = \frac{\sigma_{population}}{\sqrt{n}}$

The standard deviation of a sampling distribution is called the _standard error_.


Can also use sampling distributions to compare other parameters - proportions, regression coefficients, or standard deviations, which also follow the central limit theorem.


## T-Distribution

[They talk about T-dist closer to end of this vid](https://www.youtube.com/watch?v=yDEvXB6ApWc&list=PL8dPuuaLjXtNM_Y-bUAhblSAdWRnmBUcr&index=21)

Continous, unimodal probability distribution that's useful to represent sampling distributions. It changes its shape based on how much information there is. It's got thicker tails for smaller sample sizes (as there is less information) - making the distribution less "tight" around the mean - showing that we're more uncertain. For bigger and bigger sample sizes the distribution gets thinner tails, and eventually becomes the same as the normal distribution. Usually, for tests about means sample sizes >= 30, it's usually the same as normal. For proportions, need at least 10 instances from each group in the proportion!

## Hypothesis Testing Applied

Cool definition from [https://youtu.be/bf3egy7TQ2Q?list=PL8dPuuaLjXtNM_Y-bUAhblSAdWRnmBUcr&t=281](here):

<u>Null Hypothesis Significance Testing</u> is a form of the reductio ad absurdum argument which tries to discredit an idea by assuming the idea is true, and then showing that if you make that assumption, something contradictory happens.

(Just some examples of reductio ad absurdum outside of stats: The Earth cannot be flat; otherwise, we would find people falling off the edge. There is no smallest positive rational number because, if there were, then it could be divided by two to get a smaller one. It's like proof by contradiction.)

### P Value

A P-Value answers the question of how rare your data is, by telling you the probability of getting data that's extreme as the data you observed, if the null hypothesis were true. P-value tells you is how extreme your sample is assuming that the null is true.

P-value = 0.1, then sample is in the top 10% most extreme samples we'd expect to see based on the distribution of sample means.

A 2-sided p-value will tell us the probability of getting a sample mean as extreme (on either side) as the one we've seen, under the null hypothesis. A 1 sided will tell us the probability of getting a sample mean as high (or low) as the one we've seen, under the null hypothesis.

NB: On Common misinterpretation of the a p-value is that it can telly ou the probability that the null hypothesis is true - [IT CAN'T!](https://youtu.be/PPD8lER8ju4?list=PL8dPuuaLjXtNM_Y-bUAhblSAdWRnmBUcr&t=234)
Ronald Fisher said: "In general, tests of significance are based on hypothetical probabilities calculated from their null hypotheses. They do no generally lead to any probability statements about the real word, but to a rational and well-defines measure of reluctance to the acceptance of the hypotheses they test."
So, getting a p-value of 0.1 doesn't mean there is a 10% chance the null hypothesis is true. 

If P-Value is not lower than alpha, then we fail to reject $H_0$. Notice that we say fail to rehect, not accept. Null hypothesis testing doesn't allow us to "accept" or provide evidence that the null is true, instead we've only failed to provide evidence that it's false. Abscense of evidence is not the evidence of abscense. Failint to reject the null doesn't mean thtere isn't an effect or relationship, just means we didn't get enough evidence to say there definitely is one.



### Z - test vs T - test

Z - tests commonly done when population standard deviation is known. T - tests commonly done when population standard deviation is unknown. We use t - distribution because the unknown population standard deviation is now a value we must estimate.



### One Sample Z - Test

[From here](https://www.youtube.com/watch?v=BWJRsY-G8u0).

In the population the average IQ is 100 with a s.d. of 15. Give a sample of 30 participants a drug which either makes them smarter, stupider or has no effect. The sample of 30 people has a mean of 140. Did the medication affect intelligence?

Note here that the population standard deviation and mean is given. Very importantly, we're testing a mean, the sample mean. The question we're basically asking is how probable is it that we saw a sample mean of 140? Now, recall, from the above, sample means are distributed normally. And they will be centered around the population mean. And that the sample standard deviation is related to the population standard deviation. But remember, it's actually the sample mean standard deviation, which gets thinner as we get more observations.

So, the normal distribution in the one sample Z-test is actually the distribution of sample means, under the null hypothesis. Mkay?

$Z = \frac{\bar{x} - \mu}{\frac{\sigma}{\sqrt{n}}} = 14.60$

Let's take an $\alpha = 0.05$ (which corresponds to Type I error happening 5% of the time - Type I error is the probability of rejecting Null Hypothesis even though the Null is correct). This is a 2 - tailed test, so rehect if Z less than -1.96 or greater than 1.96 :) HERE WE REJECT NULL HYPOTHESIS.

### One Sample Z- Test for proportions

[From here](https://www.youtube.com/watch?v=dH6igFVoCAw&list=PL568547ACA9211CCA&index=61)

Survery claims 9/10 docs recommend aspirin. Survery, random sample, of 100 docs. 82 claim they recommend aspirin. At alpha = 0.05, do we believe?

$H_0$: $p = 0.9$
$H_q$: $p \neq 0.9$

The test statistic:

$Z = \frac{\hat{p} - p_0}{\sqrt{\frac{p_0(1-p_0)}{n}}} = -2.667$

$\hat{p} = 0.82$
$p_0 = 0.9$
$n = 100$

I think this comes from Bernoulli... Mean of Bernoulli is $p_0 = 0.9$ and variance is $p_0(1 - p_0) = 0.9 \times 0.1 = 0.09$

So, really I think you can derive the test for proportions from the initial equation: 

$Z = \frac{\bar{x} - \mu}{\frac{\sigma}{\sqrt{n}}}$

Except now, the population mean under the Null Hypothesis is $p_0$, the sample mean is $0.82$, the population variance, under the Null is $0.9(1 - 0.1)$, which makes the sample mean's s.d. $\sqrt{\frac{p_0(1-p_0)}{n}}$. The thing about the Bernoulli is that if you've got the mean, you've got the variance.

Anyway, -2.667 < -1.96 so reject $H_0$.

### One sample t - test

Population, average IQ = 100. New medication, either has negative, positive or no effect on intelligence. Sample of 30 participants have taken med, end up with a mean of 140 and s.d. of 20. Did medication affect intelligence?

$H_0; \mu = 100$
$H_1; \mu \neq 100$

Degrees of freedom= $n - 1 = 30 - 1 = 29$.

Critical values: -2.0452 & 2.0452.

$t = \frac{\bar{x} - \mu}{s / \sqrt{n}} = 10.96$, reject $H_0$.


## Confidence Intervals

The mean of sample distribution is the center a CI (see definition below). The way we choose the confidence range is related to the distribution of sample means (depends on standard deviation of sample means.).

The 95% in a 95% confidence interva; tells us that if we calculated a confidence interval from 100 different samples, about 95% of them would contain the true population mean. Think about it, if you got 2 different samples of 40 people, both of those would have different sample means and sample standard deviations, and thus different confidence intervals. It is possible the CI we've created doesn't contain the true mean.


## Bayes' Theorem

$P(A|B) = \frac{P(B|A)P(A)}{P(B)} = \frac{P(A \cap B)}{P(B)}$

Updating beliefs.

Suppose your sister tells you she has a friend. At this point, 50-50 chance that it's male / female.
Suppose your sister also tells you that the friend has breast cancer.
Then $P(male | BREAST CANCER) = \frac{P(BREAST CANCER | male)P(male)}{P(BREAST CANCER)}$
Suppose you know $P(BREAST CANCER | male) and P(BREAST CANCER)$ from gov't statistics, then you can calc:
$P(male | BREAST CANCER) = \frac{P(0.001)P(0.5)}{P(0.063)} = 0.79%$, so pretty low.

So, in the above, we updated our belief based on new information. Rearrange the formula slightly to see this more clearly:
$P(A|B) = \frac{P(B|A)}{P(B)}P(A) $
$P(male | BREAST CANCER) = \frac{P(0.001)}{P(0.063)}P(A)$

Updating beliefs is core to Bayesian stats.

Bayes' Factor... Say you're trying to figure out if someone is a star wars fan.
Say your _prior_ belief is that 60% of people are Star Wars fans. ie, $P(fan) = 0.6$
Suppose also, that the probabilty of watching a new star wars movie that's just come out is $P(movie | fan) = 0.99$, as fans would rush to watch it. But suppose also that the probability of watching the movie given that a person is not a fan is $P(movie | not fan) = 0.5$. This is because some people are just curious about it, (like me), or dragged by friends / family. So $\frac{P(movie | fan)}{P(movie | not fan)} =  \frac{0.99}{0.5} = 1.98$ is the odds that a person is a fan, if they tell you that they've watched the movie. It's what we've learned about they hypothesis from the data (in this case the data is someone telling you they've watched the movie). We cam multiply our _prior_ with this factor to get our posterior.

We can repeat this process; make the new posterior a prior for updating beliefs. In the example before, let's say that there are 60% star wars fans and 40% not - this is your prior. So, odds you're a fan is $\frac{0.6}{0.4} = 1.5$. But we can update this with the data: $\frac{0.6}{0.4} \times \frac{0.99}{0.5} = 1.5 \times 1.98 = 2.97$; so the odds have increased.

### Bayesian Hypothesis Testing

Compare the probabilities of the 2 hypotheses: from the above, compare the probabilities of the probability that a person is a fan, given they'd seen the movie. So, compare $P(fan | seenmovie)$ and $P(not fan | seenmovie)$.

So look that the ratio: $\frac{P(fan | seenmovie)}{P(not fan | seenmovie)}$. We look at the ratio of 2 instances of Bayes' thm; since both the numerator and the denominator are posterior probs.

### Criticism of Bayes'

In Bayesian hypothesis testing (also in just applying Bayes' thm), the result is influenced by your prior _belief_. If you throw in a different number for your belief, then you get different results. 

One of the main uses of statistics is science whis is supposed to be relatively "objective" and not influenced by opinion, yet here's a method that included beliefs in its calculation.

In some research, the researchers publish the Bayes Factor, for just this reason - it's free of the prior. You could then use their calculated Bayes factor to update your own prior.

### Bayesian Inference

Using Bayesian methods to analyze questions allows us to "inject" _prior beliefs_ into the analysis.

Using Null Hypothesis significance testing, we can ask whether means of groups are likely different, or not (reject, or fail to reject null hypothesis - recall, you can only fail to reject the null hypothesis, that is, you don't state that the means are the same, just that you haven't seen enough evidence to show they aren't). With Bayesian methods, you can assign a probability to the question of whether or not the means are different. It's not just a binary outcome... (I think...)

Bayesian A/B Testing: https://www.countbayesie.com/blog/2015/4/25/bayesian-ab-testing
^^ In the above he also mentions why to use instead of using t-tests for differences of means.
"Classical statistics tells us Significance, but what we're really after is Magnitude!" - sayeth Count Bayesie.

# Definitions

- <u>Bayes' Factor</u> - Represents the amount of info we've learned about our hypotheses from the data.
- <u>Central Limit Theorem</u> - The distribution of sample means for an independent random variable will get closer and closer to a normal distribution as the size of the sample gets bigger and bigger, even if the original population distribution isn't normal. Think of independet die rolls, like in [here](https://youtu.be/rBjft49MAO8?list=PL8dPuuaLjXtNM_Y-bUAhblSAdWRnmBUcr&t=349). Twenty dice are 20 independent random variables.
- <u>Confidence Interval</u> - An <b>estimated range of values</b> that seem reasonable based on what we've observed. Its center is still the <b>sample mean</b>, but we've got some room  on either side for out uncertainty.
- <u>Hypthesis Testing</u> - 
- <u>Inference</u> - the process of drawing conclusions about population parameters based on a sample taken from the population.
- <u>P-Value</u> - probability of obtaining test results at least as extreme as the results actually observed, assuming that the null hypothesis is correct.
- <u>Point Estimate </u> - a single value given as an estimate of a parameter of a population. Like the mean.
- <u>Sampling distribution</u> - the probability distribution of a sample statistic (like the mean, or variance).
- <u>Standard Error</u> - of a statistic (usually an estimate of a parameter - like the mean) is the standard deviation of its sampling distribution. For example, Standard error of the mean is $\sigma_{\bar{x}} = \frac{\sigma}{\sqrt{n}}$
- <u>Statistical Power </u> - $1 - \beta$ - chance of detecting an effect if there is one.
- <u>Test Statistic </u> - Allow us to quantify how close things are to our expectations or theories
- <u>Type I error</u> - $\alpha$ - probability of saying null hypothesis is wrong, when it is correct
- <u>Type II error</u> - $\beta$ - probability of not saying the null hypothesis is wrong, even though it is wrong. Note, you're not actually saying it's true, just that it's not wrong. 
- <u>Z-Distrubtion</u> - Standardized normal distribution
- <u>Z-Score</u> - A Z-score is a numerical measurement that describes a value's relationship to the mean of a group of values. Z-score is measured in terms of standard deviations from the mean. If a Z-score is 0, it indicates that the data point's score is identical to the mean score. A Z-score of 1.0 would indicate a value that is one standard deviation from the mean. __Z scores allow us to compare things that are not on the same scale as long as they are normally distributed__.


