---
title: An R Pipeline for XGBoost (and a discussion about hyperparameters)
author: 'Orry Messer'
date: '2019-10-04'
slug: an-r-pipeline-for-xgboost-and-a-discussion-about-hyperparameters
draft: true
categories: [data-science]
tags:
  - R
  - data-science
  - xgboost
output:
  blogdown::html_page:
      number_sections: true
      toc: TRUE
---

```{r, results='hide'}
library(Matrix)
library(xgboost)
library(ggplot2)
library(readr)
library(dplyr)
```


# Introduction

[XGBoost](https://xgboost.readthedocs.io/en/latest/) is an implementation of a machine learning technique known as [gradient boosting](https://en.wikipedia.org/wiki/Gradient_boosting).
In this blog post, we discuss what XGBoost is, and demonstrate a pipeline for working with it in R.

# XGBoost - An Implementation of Gradient Boosting

Gradient boosting is part of a class of machine learning techniques known as ensemble methods. An ensemble method leverages the output of many <u>weak learners</u> in order to make a prediction. Typically, these weak learners are implemented as decision trees. While each individual weak learner might not get the answer right, on average, their combined answers should be pretty decent. What makes gradient boosting different from another popular ensemble method - [Random Forest](https://en.wikipedia.org/wiki/Random_forest), is that the construction of the weak learners depend on the previously constructed learners. 

In gradient boosting, each weak learner is chosen iteratively in a greedy manner, so as to minimize the loss function.

XGBoost is a highly optimized implementation of gradient boosting. The original paper describing XGBoost can be found [here](https://arxiv.org/pdf/1603.02754.pdf). Although XGBoost is written in C++, it can be interfaced from R using the `xgboost` package.

To install the package:

`install.packages("xgboost")`


# Load And Explore The Data

We will use the [Titanic Dataset](https://www.kaggle.com/c/titanic/data). Basically, we try predict whether a passenger survived or not (so this is a binary classification problem).

Let's load up the data:

```{r, results='hide'}

titanic_train <- read_csv("./xg_boost_data/train.csv")
titanic_test <- read_csv("./xg_boost_data//test.csv")

```

For the sake of brevity, I'll only keep some of the features:

- Pclass
- Sex
- Age
- Embarked

Let's have a look at our data after discarding a few features:

```{r}

titanic_train <- titanic_train %>%
  select(Survived,
         Pclass,
         Sex,
         Age,
         Embarked)

titanic_test <- titanic_test %>%
  select(Pclass,
         Sex,
         Age,
         Embarked)

str(titanic_train)

```
XGBoost will only take numeric data as input. Let's convert our character features to factors, and one-hot encode.

```{r}



# sparse.model.matrix() will drop rows with NA's (https://stackoverflow.com/questions/29732720/sparse-model-matrix-loses-rows-in-r)
summary(titanic_train) # seems like there are 177 NA's in the Age variable, and 2 in the Embarked variable
summary(titanic_test) # 68 NA's in age var

# We don't want to drop rows. So let's replace NA's with a sentinal value; how about -999?
titanic_train[is.na(titanic_train)] <- -999
titanic_test[is.na(titanic_test)] <- -999

titanic_train$Sex <- as.factor(titanic_train$Sex)
titanic_train$Embarked <- as.factor(titanic_train$Embarked)
titanic_train$Pclass <- as.factor(titanic_train$Pclass) # Could be ordinal, but leaving it is strict categorical
titanic_test$Sex <- as.factor(titanic_test$Sex)
titanic_test$Embarked <- as.factor(titanic_test$Embarked)
titanic_test$Pclass <- as.factor(titanic_test$Pclass) # Could be ordinal, but leaving it is strict categorical

titanic_train_sparse <- sparse.model.matrix(Survived~., data = titanic_train)[,-1]
# Recall, xgboost takes advantage of the sparsity. Sparsity can be induced from 1-hot encoding.
class(titanic_train_sparse)
```

The data are in the format of a __dgCMatrix__ class - this is the Matrix package's implementation of sparse matrix.

Let's have a look at the structure of the data a little closer:

```{r}
str(titanic_train_sparse)
```

We can check this directly:

```{r}
dim(titanic_train_sparse)
```

The names are the features are given by titanic_train_sparse@Dimnames:

```{r}
head(titanic_train_sparse@Dimnames[[2]])
```

You can convert this data into a dataframe, thusly:

```{r, results='hide'}
train_data_as_df <- as.data.frame(as.matrix(titanic_train_sparse))
```

# 3 Hyperparameters

This is a vast topic. Without going into too much depth, I'll outline some of the more commonly used hyperparameters:

```{r}
# Full reference: https://xgboost.readthedocs.io/en/latest/parameter.html
#### Tree booster params...####
# eta:       						default = 0.3
#            						learning rate / shrinkage. Scales the contribution of each try by a factor of 0 < eta < 1
# gamma:     						default = 0
#            						minimum loss reduction needed to make another partition in a given tree.
#            						larger the value, the more conservative the tree will be (as it will need to make a bigger reduction to split)
#            						So, conservative in the sense of willingness to split.
# max_depth: 						default = 6
#            						max depth of each tree...
# subsample: 						1 (ie, no subsampling)
#            						fraction of training samples to use in each "boosting iteration"
# colsample_bytree:     default = 1 (ie, no sampling)
#                       Fraction of columns to be used when constructing each tree. This is an idea used in RandomForests
# min_child_weight:     default = 1
#                       This is the minimum number of instances that have to been in a node. It's a regularization parameter
#                       So, if it's set to 10, each leaf has to have at least 10 instances assigned to it.
#                       The higher the value, the more conservative the tree will be.



```

(I've left it as commented code, as I like to past this into my scripts as a quick reference.)

Let's create the hyper-parameters list:

```{r, results='hide'}
params_booster <- list(
  booster = 'gbtree', # Possible to also have linear boosters as your weak learners.
  eta = 1, 
  gamma = 0,
  max.depth = 2, 
  subsample = 1, 
  colsample_bytree = 1,
  min_child_weight = 1, 
  objective = "binary:logistic"
)

bstSparse <- xgboost(data = titanic_train_sparse, 
                     label = titanic_train$Survived, 
                     nrounds = 100,  
                     params = params_booster)

```

The xgb.train() and xgboost() functions are used to train the boosting model, and both return an object of class xgb.Booster.
Before we do that, let's first use xgb.cv() to get some understanding of our performance before we evaluate against our final hold our test set. Important to not that xgb.cv() returns an object of type xgb.cv.synchronous, not xgb.Booster. So you won't be able to call functions like xgb.importance() on it, as xgb.importance() takes object of class xgb.Booster __not__ xgb.cv.synchronous.


```{r, results='hide'}
# NB: keep in mind xgb.cv() is used to select the correct hyperparams.
# Once you have them, train using xgb.train() or xgboost() to get the final model.

bst.cv <- xgb.cv(data = titanic_train_sparse, 
              label = titanic_train$Survived, 
              params = params_booster,
              nrounds = 300, 
              nfold = 5,
              print_every_n = 20,
              verbose = 2)



```

Note, we can also implement early-stopping: early_stopping_rounds = 3, so that if there has been no improvement in test accuracy for a specified number of rounds, the algorithm stops.

```{r}
res_df <- data.frame(tr = bst.cv$evaluation_log$train_error_mean, 
                     val = bst.cv$evaluation_log$test_error_mean,
                     iter = bst.cv$evaluation_log$iter)

g <- ggplot(res_df, aes(x=iter)) +        # Look @ it overfit.
  geom_line(aes(y=tr), color = "blue") +
  geom_line(aes(y=val), colour = "green")

g

```

